---
title: "EDA"
author: "Group 06"
date: "January 17, 2019"
output:
  html_document: default
  pdf_document: default
---
# Step_1: clean the environment and import the library
* The loaded librariesin the code are:
1. tidyverse
2. here 
3. DataExplorer
```{r setup, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Clear environment
rm(list = ls(all = TRUE))

# check if package is installed, if not installed then load
if(!require(here)){
  install.packages("here")
}
library(here)

if(!require(tidyverse)){
  install.packages("tidyverse")
}
library(tidyverse)

if(!require(DataExplorer)){
  install.packages("DataExplorer")
}
library(DataExplorer)
```

# Step_2: import the data
* the two imported datasets will be merged in the dataframe of "base" on the variable of "SERIALNO" 
1. import the ss16pal.csv
```{r echo=TRUE, message=FALSE}
#load the persional data to variable Base
base <- read_csv(here('raw_data','ss16pal.csv'))%>% 
  select(SERIALNO, #Serial Number to link House Hold and Personal data
         AGEP,  # Age: 0#99
         CIT,   # Citizenship status
         COW,   # Class of worker
         DDRS,  # Self#care difficulty
         DEAR,  # Hearing difficulty
         DEYE,  # Vision difficulty
         DOUT,  # Independent living difficulty
         DPHY,  # Ambulatory difficulty
         DRAT,  # Veteran service connected disability rating (percentage)
         DRATX, # Veteran service connected disability rating (checkbox)
         DREM,  # Cognitive difficulty
         ENG,   # Ability to speak English
         JWMNP, # Travel time to work
         JWRIP, # Vehicle occupancy
         JWTR,  # Means of transportation to work
         MAR,   # Marital status
         MIL,   # Military service
         NWLK,  # Looking for work
         SCHL,  # Educational attainment
         SEMP,  # Self-employment income past 12 months (signed)
         SEX,   # Sex
         WAGP,  # Wages or salary income past 12 months
         ESR,   # Employment status recode
         HISP,  # Recoded detailed Hispanic origin
         NATIVITY,  # Nativity
         PERNP, # Total person's earnings
         RAC1P, # Recoded detailed race code
         RAC2P, # Recoded detailed race code
         RAC3P, # Recoded detailed race code
         RACAIAN, # American Indian and Alaska Native recode
         RACASN,  # Asian recode
         RACBLK,  # Black or African American recode
         RACNH, # Native Hawaiian recode
         RACWHT,  # White recode
         WAOB,  # World area of birth
         WKW)   #Weeks worked during past 12 months
```

2. import the ss16hal.csv
```{r echo=TRUE, message=FALSE}
#Link the house hould data
househould <- read_csv(here('raw_data','ss16hal.csv'))%>% 
  select(SERIALNO, #Serial Number to link House Hold and Personal data
         TYPE, # Type of house
         ACCESS, # Access to the Internet
         BROADBND, # Cellular data plan for a smartphone or other mobile device
         COMPOTHX, # Other computer equipment
         HISPEED, #Broadband (high speed) Internet service such as cable, fiber optic, or DSL service
         LAPTOP, # Laptop or desktop
         OTHSVCEX, # Other Internet service
         SATELLITE, # Satellite Internet service
         SMARTPHONE, # Smartphone
         TABLET # Tablet or other portable wireless computer
  )

```

3. merge the two datasets on "SERIALNO"
```{r echo=TRUE, message=FALSE}
base <- base %>% left_join(househould) %>% select(-SERIALNO)
rm(househould)
```



# Step_3: data struction of dataframe of "base"
```{r message=FALSE}
str(base)
```


# Step_4: convert the data
* the categorical variables are in the 
1. convert the following variables into factors
```{r}
base <- base %>% mutate(CIT = as.factor(CIT),
                        COW = as.factor(COW),
                        DDRS = as.factor(DDRS),
                        DEAR = as.factor(DEAR),
                        DEYE = as.factor(DEYE),
                        DOUT = as.factor(DOUT),
                        DPHY = as.factor(DPHY),
                        DRAT = as.factor(DRAT),
                        DRATX = as.factor(DRATX),
                        DREM = as.factor(DREM),
                        ENG = as.factor(ENG),
                        JWRIP = as.factor(JWRIP),
                        JWTR = as.factor(JWTR),
                        MAR = as.factor(MAR),
                        MIL = as.factor(MIL),
                        NWLK = as.factor(NWLK),
                        SCHL = as.factor(SCHL),
                        SEX = as.factor(SEX),
                        ESR = as.factor(ESR),
                        HISP= as.factor(HISP),
                        RAC1P= as.factor(RAC1P),
                        RAC2P= as.factor(RAC2P),
                        RAC3P= as.factor(RAC3P),
                        RACAIAN = as.factor(RACAIAN),
                        RACASN = as.factor(RACASN),
                        RACBLK = as.factor(RACBLK),
                        RACNH = as.factor(RACNH),
                        RACWHT = as.factor(RACWHT),
                        WAOB = as.factor(WAOB),
                        WKW = as.factor(WKW),
                        TYPE = as.factor(TYPE),
                        ACCESS = as.factor(ACCESS),
                        BROADBND = as.factor(BROADBND),
                        COMPOTHX = as.factor(COMPOTHX),
                        HISPEED = as.factor(HISPEED),
                        LAPTOP = as.factor(LAPTOP),
                        OTHSVCEX = as.factor(OTHSVCEX),
                        SATELLITE = as.factor(SATELLITE),
                        SMARTPHONE = as.factor(SMARTPHONE),
                        TABLET = as.factor(TABLET)
                        )

```
2. convert the following variables into numeric
```{r}
base <- base %>% mutate(AGEP = as.numeric(AGEP),
                        JWMNP = as.numeric(JWMNP),
                        SEMP= as.numeric(SEMP),
                        WAGP= as.numeric(WAGP),
                        PERNP= as.numeric(PERNP)
                        )
```

```{r}
base2 <- base$WAOB
```


# step_6: EDA
## Step_6.1 EDA - Data Structure

```{r}
str(base)
```

## Step_6.2 EDA - Data Summary
```{r}
summary(base)
```

