---
title: "The Impact of Technology Access on Earnigs"
author: "Group 06"
date: "January 17, 2019"
output:
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document: default
  word_document: default
---
With a past of racial inequality in the southern states it would be interesting to examine how race/ethnicity is related to access to the internet and other technologies and how not having or having access could effect a persons earnings. This can be explored via the following research questions:
In the state of Alabama, how to earnings vary by access to the internet/technology?
Does the internet/technology premium vary by race and ethnicity?

# Step_1: System Setup
Before starting the program we need to clean the environment and import the library.
The loaded librariesin the code are:
  1. tidyverse
  2. here 
  3. DataExplorer
```{r setup, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Clear environment
rm(list = ls(all = TRUE))

# check if package is installed, if not installed then load
if(!require(here)){
  install.packages("here")
}
library(here)

if(!require(tidyverse)){
  install.packages("tidyverse")
}
library(tidyverse)

if(!require(DataExplorer)){
  install.packages("DataExplorer")
}
library(DataExplorer)
source(here("code","function_convert_CSV_to_vector.R"))
```

# Step_2: Import Data

## Step_2.1: Import personal income Data

Here we will import the raw data of personal income (ss16pal) into the dataframe as "base"
```{r import_data, message=FALSE}
#load the persional data to variable Base
base <- read_csv(here('raw_data','ss16pal.csv'))%>% 
  select(SERIALNO, #Serial Number to link House Hold and Personal data
         AGEP,  # Age: 0#99
         CIT,   # Citizenship status
         COW,   # Class of worker
         DDRS,  # Self#care difficulty
         DEAR,  # Hearing difficulty
         DEYE,  # Vision difficulty
         DOUT,  # Independent living difficulty
         DPHY,  # Ambulatory difficulty
         DRAT,  # Veteran service connected disability rating (percentage)
         DRATX, # Veteran service connected disability rating (checkbox)
         DREM,  # Cognitive difficulty
         ENG,   # Ability to speak English
         JWMNP, # Travel time to work
         JWRIP, # Vehicle occupancy
         JWTR,  # Means of transportation to work
         MAR,   # Marital status
         MIL,   # Military service
         NWLK,  # Looking for work
         SCHL,  # Educational attainment
         SEMP,  # Self-employment income past 12 months (signed)
         SEX,   # Sex
         WAGP,  # Wages or salary income past 12 months
         ESR,   # Employment status recode
         HISP,  # Recoded detailed Hispanic origin
         NATIVITY,  # Nativity
         PERNP, # Total person's earnings
         RAC1P, # Recoded detailed race code
         RACAIAN, # American Indian and Alaska Native recode
         RACASN,  # Asian recode
         RACBLK,  # Black or African American recode
         RACNH, # Native Hawaiian recode
         RACWHT,  # White recode
         WAOB,  # World area of birth
         WKW)   #Weeks worked during past 12 months
```

## Step_2.2: Import household Data

We need to import the household level information (ss16hal) about technology access and merge the data into "base"

```{r household_info, message=FALSE}
#Link the house hould data
househould <- read_csv(here('raw_data','ss16hal.csv'))%>% 
  select(SERIALNO, #Serial Number to link House Hold and Personal data
         TYPE, # Type of house
         ACCESS, # Access to the Internet
         BROADBND, # Cellular data plan for a smartphone or other mobile device
         COMPOTHX, # Other computer equipment
         HISPEED, #Broadband (high speed) Internet service such as cable, fiber optic, or DSL service
         LAPTOP, # Laptop or desktop
         OTHSVCEX, # Other Internet service
         SATELLITE, # Satellite Internet service
         SMARTPHONE, # Smartphone
         TABLET # Tablet or other portable wireless computer
  )
base <- base %>% left_join(househould) %>% select(-SERIALNO)
rm(househould)
```

# Step_3: First look of data structure.
```{r first_look}
str(base)
```


# Step_4: Converting data

Convert the categorical data into factor type and numetical data into numeric type. The factorial data in this dataframe will keep its representative code instead of the category name to save the memory space. 
```{r converting_data, message=FALSE}
#conver the catagorical variable type
base <- base %>% mutate(CIT = as.factor(CIT),
                        COW = as.factor(COW),
                        DDRS = as.factor(DDRS),
                        DEAR = as.factor(DEAR),
                        DEYE = as.factor(DEYE),
                        DOUT = as.factor(DOUT),
                        DPHY = as.factor(DPHY),
                        DRAT = as.factor(DRAT),
                        DRATX = as.factor(DRATX),
                        DREM = as.factor(DREM),
                        ENG = as.factor(ENG),
                        JWRIP = as.factor(JWRIP),
                        JWTR = as.factor(JWTR),
                        MAR = as.factor(MAR),
                        MIL = as.factor(MIL),
                        NWLK = as.factor(NWLK),
                        SCHL = as.factor(SCHL),
                        SEX = as.factor(SEX),
                        ESR = as.factor(ESR),
                        HISP= as.factor(HISP),
                        RAC1P= as.factor(RAC1P),
                        RACAIAN = as.factor(RACAIAN),
                        RACASN = as.factor(RACASN),
                        RACBLK = as.factor(RACBLK),
                        RACNH = as.factor(RACNH),
                        RACWHT = as.factor(RACWHT),
                        WAOB = as.factor(WAOB),
                        WKW = as.factor(WKW),
                        TYPE = as.factor(TYPE),
                        ACCESS = as.factor(ACCESS),
                        BROADBND = as.factor(BROADBND),
                        COMPOTHX = as.factor(COMPOTHX),
                        HISPEED = as.factor(HISPEED),
                        LAPTOP = as.factor(LAPTOP),
                        OTHSVCEX = as.factor(OTHSVCEX),
                        SATELLITE = as.factor(SATELLITE),
                        SMARTPHONE = as.factor(SMARTPHONE),
                        TABLET = as.factor(TABLET)
                        )

#conver the numeric variable type
base <- base %>% mutate(AGEP = as.numeric(AGEP),
                        JWMNP = as.numeric(JWMNP),
                        SEMP= as.numeric(SEMP),
                        WAGP= as.numeric(WAGP),
                        PERNP= as.numeric(PERNP)
                        )
```

Replace the factor level to the choices in the survery question for easy understanding.
```{r, message=FALSE}
levels(base$ACCESS) <- convertCSV2Factor("ACCESS")
levels(base$BROADBND) <- convertCSV2Factor("BROADBND")
levels(base$CIT) <- convertCSV2Factor("CIT")
levels(base$COMPOTHX) <- convertCSV2Factor("COMPOTHX")
levels(base$COW) <- convertCSV2Factor("COW")
levels(base$DDRS) <- convertCSV2Factor("DDRS")
levels(base$DEAR) <- convertCSV2Factor("DEAR")
levels(base$DEYE) <- convertCSV2Factor("DEYE")
levels(base$DOUT) <- convertCSV2Factor("DOUT")
levels(base$DOUT) <- convertCSV2Factor("DOUT")
levels(base$DPHY) <- convertCSV2Factor("DPHY")
levels(base$DRAT) <- convertCSV2Factor("DRAT")
levels(base$DRATX) <- convertCSV2Factor("DRATX")
levels(base$DREM) <- convertCSV2Factor("DREM")
levels(base$ENG) <- convertCSV2Factor("ENG")
levels(base$ESR) <- convertCSV2Factor("ESR")
levels(base$HISP) <- convertCSV2Factor("HISP")
levels(base$HISPEED) <- convertCSV2Factor("HISPEED")
levels(base$JWRIP) <- convertCSV2Factor("JWRIP")
levels(base$JWTR) <- convertCSV2Factor("JWTR")
levels(base$LAPTOP) <- convertCSV2Factor("LAPTOP")
levels(base$MAR) <- convertCSV2Factor("MAR")
levels(base$MIL) <- convertCSV2Factor("MIL")
levels(base$NWLK) <- convertCSV2Factor("NWLK")
levels(base$OTHSVCEX) <- convertCSV2Factor("OTHSVCEX")
levels(base$RAC1P) <- convertCSV2Factor("RAC1P")
levels(base$RACAIAN) <- convertCSV2Factor("RACAIAN")
levels(base$RACASN) <- convertCSV2Factor("RACASN")
levels(base$RACBLK) <- convertCSV2Factor("RACBLK")
levels(base$RACNH) <- convertCSV2Factor("RACNH")
levels(base$RACWHT) <- convertCSV2Factor("RACWHT")
levels(base$SATELLITE) <- convertCSV2Factor("SATELLITE")
levels(base$SCHL) <- convertCSV2Factor("SCHL")
levels(base$SEX) <- convertCSV2Factor("SEX")
levels(base$SMARTPHONE) <- convertCSV2Factor("SMARTPHONE")
levels(base$TABLET) <- convertCSV2Factor("TABLET")
levels(base$WAOB) <- convertCSV2Factor("WAOB")
levels(base$WKW) <- convertCSV2Factor("WKW")
```
==================

# Step_5: Exploratory data analysis - EDA

## Step_5.1 EDA - Data Structure

```{r}
str(base)
```

## Step_5.2 EDA - Data Summary
```{r}
summary(base)
```

## Step_5.3 EDA - Age:
According to the graph and table below we see that our data includes people from 0 to 99 years old.
```{r eda_age}
base %>% select(AGEP) %>%  summary()
base %>% ggplot(aes(AGEP)) + geom_histogram()  
```

We decided to work with people from 16 to 50 years olds as a better representations of the market labor
```{r eda_age_filter}
base <- base %>%  filter(AGEP>=16 & AGEP <= 60)
base %>% select(AGEP) %>%  summary()
base %>% ggplot(aes(AGEP)) + geom_density()  
```

## Step_5.4 EDA - Labor:  
We decided to exclude people:  
  1. Working without pay in family business or farm  
  2. Unemployed and last worked 5 years ago or earlier or never worked  
```{r }
base <- base %>%  filter(as.integer(COW)<8)
```
  3. Worked less than 50 weeks in the last 12 months
```{r }
base <- base %>%  filter(as.integer(WKW)==1)
```  


## Step_5.5 EDA - Wage:  
Since the focus of our analysis is access to technology we can restrict the data by removing the rich and upper middle class (setting the max at the 3rd Quartile) because at a certain wage level we can conclude that technology becomes more about choice than access.  
```{r }
base  %>% ggplot(aes(WAGP)) + geom_histogram()
base <- base %>%  filter(WAGP<54000 & WAGP > 1000)
summary(base$WAGP)
```
Our earnings distributions after this considerations is:
```{r }
base  %>% ggplot(aes(WAGP)) + geom_density()
base  %>% ggplot(aes(y=WAGP)) + geom_boxplot()

```

## Step_5.5 EDA - Missing data:
```{r}
plot_missing(base)
```

Since the variable of "DRATX", "ENG", "DRAT" have numerous missing data that lead these three vairable as "un-representative," so this study will remove these three variables.

```{r}
base$DRAT <- NULL
base$ENG <- NULL
base$DRATX <- NULL
```


# Step_6: Exploratory data analysis after restrictions
```{r }
summary(base)
```

==================

# Step_7: Examine access to the internet and its relation to earnings.

```{r}
base %>% select(ACCESS) %>% summary()

base %>%  filter(!is.na(ACCESS)) %>% 
  ggplot(aes(WAGP,color=ACCESS)) + geom_density() + 
  labs(title="Earnings x Access to the internet",
       x ="Earnings", y = "Density") +
  scale_color_discrete(name="Internet Access",
                      labels=c("Yes- Paying", "Yes-Not Paying", "No"))

base %>%  filter(!is.na(ACCESS)) %>%
  ggplot() + geom_histogram(aes(x=WAGP,fill=ACCESS), binwidth = 10000) 
```

## Step_7.1 Using Tablet as a proxy of Technology Access

```{r }
#base %>% select(TABLET) %>% summary()
plot_bar(base$TABLET)
base %>% filter(TABLET=="Yes") %>% select(WAGP) %>% summary() #person have tablet
base %>% filter(TABLET=="No") %>% select(WAGP) %>% summary() #person doesn't have tablet

base %>%  filter(!is.na(TABLET)) %>%
  ggplot(aes(WAGP,color=TABLET)) + geom_density() + 
  labs(title="Earnings x Having Tablet",
       x ="Earnings", y = "Density") +
  scale_color_discrete(name="Person has Tablet",
                      labels=c("Yes", "No"))

base %>%  filter(!is.na(TABLET)) %>%
  ggplot() + geom_histogram(aes(x=WAGP,fill=TABLET), binwidth = 5000) 
```


## Step_7.2 Using Smartphone as a proxy of Technology Access
```{r }
base %>% select(SMARTPHONE) %>% summary()
base %>% filter(SMARTPHONE=="Yes") %>% select(WAGP) %>% summary() #person have smartphone
base %>% filter(SMARTPHONE=="No") %>% select(WAGP) %>% summary() #person doesn't have smartphone

base %>%  filter(!is.na(SMARTPHONE)) %>%
  ggplot(aes(WAGP,color=SMARTPHONE)) + geom_density() + 
  labs(title="Earnings x Having Smartphone",
       x ="Earnings", y = "Density") +
  scale_color_discrete(name="Person has Smartphone",
                      labels=c("Yes", "No"))

base %>%  filter(!is.na(SMARTPHONE)) %>%
  ggplot() + geom_histogram(aes(x=WAGP,fill=SMARTPHONE), binwidth = 5000) 
```
## Step_7.3 Using Laptop as a proxy of Technology Access
```{r }
base %>% select(LAPTOP) %>% summary()
base %>% filter(LAPTOP=="Yes") %>% select(WAGP) %>% summary() #person have Laptop
base %>% filter(LAPTOP=="No") %>% select(WAGP) %>% summary() #person doesn't have Laptop

base %>%  filter(!is.na(LAPTOP)) %>%
  ggplot(aes(WAGP,color=LAPTOP)) + geom_density() + 
  labs(title="Earnings x Having Laptop",
       x ="Earnings", y = "Density") +
  scale_color_discrete(name="Person has Laptop",
                      labels=c("Yes", "No"))

base %>%  filter(!is.na(LAPTOP)) %>%
  ggplot() + geom_histogram(aes(x=WAGP,fill=LAPTOP), binwidth = 5000)
```

## Step_7.4 Using HIGH SPEED Internet as a proxy of Technology Access
```{r }
base %>% select(HISPEED) %>% summary()
base %>% filter(HISPEED=="Yes") %>% select(WAGP) %>% summary() #person have HIGH SPEED Internet
base %>% filter(HISPEED=="No") %>% select(WAGP) %>% summary() #person doesn't have HIGH SPEED Internet

base %>%  filter(!is.na(HISPEED)) %>%
  ggplot(aes(WAGP,color=HISPEED)) + geom_density() + 
  labs(title="Earnings x Having High Speed Internet",
       x ="Earnings", y = "Density") +
  scale_color_discrete(name="Person has High Speed Internet",
                      labels=c("Yes", "No"))

base %>%  filter(!is.na(HISPEED)) %>%
  ggplot() + geom_histogram(aes(x=WAGP,fill=HISPEED), binwidth = 5000)
```

## Step_7.5 Race and access to the internet

```{r}
base %>% select(RAC1P) %>% summary()
base %>% filter(ACCESS=="No access to the Internet at this house, apartment, or mobile home") %>% select(RAC1P) %>% summary() #person doesn't have ACCESS to Internet

base %>%  filter(!is.na(ACCESS)) %>%
  ggplot() + geom_bar(aes(x = RAC1P, fill=ACCESS)) + 
  coord_flip() + 
  scale_x_discrete(labels = c('White', 'Black', 'Native American', "Multiple tribes", "Asian", 'Pacific Islander', 'Other', 'Two or more Races')) +
  theme(legend.position= c(0.62,0.85))

```

==================

# Notes on variable selection and restriction

By examining the relationship between earnings and access to technology we are able to see if an individuals earnings are at all effected by their level of access to technology. This is of particular interest when applied to the middle class and working poor. When performing variable selection and restriction, we chose to restrict the age variable to adults of working age from 16 to 60. Additionally, we chose to remove respondents who were working without pay, who are unemployed, or who have worked for less than 50 weeks.

Once we restricted the work related variables, we examined the earnings variable (WAGP). Since our analysis is focused on access to technology we removed the top quartile of earnings. This is because over a certain earnings rate technology acquisition is based more on choice than access. We also restricted the lower rate to $1000, we did this to limit the respondents who claimed to have worked for at least 50 weeks but reported $0 in earnings.

Lastly, we have a variable for access to the internet (ACCESS) but as a proxy for access to technology we have smartphones, tablets and laptops. At this point in our analysis we can determine that smartphones may not be a good proxy because it appears that almost every respondent has access to a smartphone. However, between tablets and laptops we may have wait until we start specifying our model to examine which variable to use as a proxy for access to technology.


