---
title: "Labor Market Analysis"
author: "Mark Russeff, Renato Albolea, Shuai Ma"
subtitle: Summary Findings
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---
&nbsp;
&nbsp;


__Research Questions:__ 

With a past of racial inequality in the southern states it would be interesting to examine how race/ethnicity is related to access to the internet and other technologies and how not having or having access could effect a persons earnings. This can be explored via the following research questions:

  1. In the state of Alabama, how to earnings vary by access to the internet/technology?
  2. Does the internet/technology premium vary by race and ethnicity?  
  
&nbsp;
&nbsp;

```{r setup, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Clear environment
rm(list = ls(all = TRUE))

# check if package is installed, if not installed then load
if(!require(here)){
  install.packages("here")
}
library(here)

if(!require(tidyverse)){
  install.packages("tidyverse")
}
library(tidyverse)

if(!require(DataExplorer)){
  install.packages("DataExplorer")
}
library(DataExplorer)

if(!require(lmtest)){
  install.packages("lmtest")
}
library(lmtest)

if(!require(stargazer)){
  install.packages("stargazer")
}
library(stargazer)
library(kableExtra)


source(here("code","function_convert_CSV_to_vector.R"))
```
```{r import_data, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
#load the persional data to variable Base
base <- read_csv(here('raw_data','ss16pal.csv'))%>% 
 dplyr::select(SERIALNO, #Serial Number to link House Hold and Personal data
         AGEP,  # Age: 0#99
         SEMP,  # Self-employment income past 12 months (signed)
         WAGP,  # Wages or salary income past 12 months
         HISP,  # Recoded detailed Hispanic origin
         NATIVITY,  # Nativity
         PERNP, # Total person's earnings
         RAC1P, # Recoded detailed race code
         RACAIAN, # American Indian and Alaska Native recode
         RACASN,  # Asian recode
         RACBLK,  # Black or African American recode
         RACNH, # Native Hawaiian recode
         RACWHT,  # White recode
         SCHL,   # Educational attainment
         SEX,     # SEX
         WAOB,  # World area of birth
         WKW)   #Weeks worked during past 12 months
```
```{r household_info, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
#Link the house hould data
househould <- read_csv(here('raw_data','ss16hal.csv'))%>% 
  dplyr::select(SERIALNO, #Serial Number to link House Hold and Personal data
        ACCESS # Access to the Internet
  )
base <- base %>% left_join(househould) %>% dplyr::select(-SERIALNO)
rm(househould)
```
```{r converting_data, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
#conver the catagorical variable type
base <- base %>% mutate(HISP= as.factor(HISP),
                        RAC1P= as.factor(RAC1P),
                        RACAIAN = as.factor(RACAIAN),
                        RACASN = as.factor(RACASN),
                        RACBLK = as.factor(RACBLK),
                        RACNH = as.factor(RACNH),
                        RACWHT = as.factor(RACWHT),
                        SCHL = as.factor(SCHL),
                        SEX = as.factor(SEX),
                        WAOB = as.factor(WAOB),
                        WKW = as.factor(WKW),
                        ACCESS = as.factor(ACCESS),
                        NATIVITY = as.factor(NATIVITY)
                        )

#conver the numeric variable type
base <- base %>% mutate(AGEP = as.numeric(AGEP),
                        SEMP= as.numeric(SEMP),
                        WAGP= as.numeric(WAGP),
                        PERNP= as.numeric(PERNP)
                        )
levels(base$ACCESS) <- convertCSV2Factor("ACCESS")
levels(base$HISP) <- convertCSV2Factor("HISP")
levels(base$RAC1P) <- convertCSV2Factor("RAC1P")
levels(base$RACAIAN) <- convertCSV2Factor("RACAIAN")
levels(base$RACASN) <- convertCSV2Factor("RACASN")
levels(base$RACBLK) <- convertCSV2Factor("RACBLK")
levels(base$RACNH) <- convertCSV2Factor("RACNH")
levels(base$RACWHT) <- convertCSV2Factor("RACWHT")
levels(base$SCHL) <- convertCSV2Factor("SCHL")
levels(base$SEX) <- convertCSV2Factor("SEX")
levels(base$WKW) <- convertCSV2Factor("WKW")
levels(base$WAOB) <- convertCSV2Factor("WAOB")
levels(base$NATIVITY) <- convertCSV2Factor("NATIVITY")
```
```{r, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
base <- base %>% 
        mutate( age_group = case_when(
                            AGEP < 14 ~ "Other ages (exclude)",
                            AGEP >= 14 & AGEP <= 62 ~ "Age 14-62 (include)",
                            AGEP > 62 ~ "Other ages (exclude)"
        ))
base$age_group <- as.factor(base$age_group)

base <- base %>%  filter(AGEP>=14 & AGEP <= 62)
base$age_group <- NULL

base <- base %>%  filter(WAGP > 0)

base <- base %>% 
         filter(WKW %in% c("50 to 52 weeks worked during past 12 months", "48 to 49 weeks worked during past 12 months", "40 to 47 weeks worked during past 12 months"))
```
&nbsp;
&nbsp;

__Table of Descriptive Statistics for Free Internet Access__


```{r, results='asis', echo=FALSE, error=FALSE, warning=FALSE }

base_access <- base %>%
  filter(ACCESS == "Free Access") %>%
  select(AGEP, WAGP)
base_no <- base %>%
  filter(ACCESS == "No Access") %>%
  select(AGEP, WAGP)

stargazer(as.data.frame(base_access), title = "Free Access", header = FALSE, summary = TRUE)
```
&nbsp;
&nbsp;

__Table of Descriptive Statistics for No Internet Access__


```{r, results='asis', echo=FALSE, error=FALSE, warning=FALSE }
stargazer(as.data.frame(base_no), title = "No Access", header = FALSE, summary = TRUE)
```

```{r, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
base_model <- base %>%  
  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>%
  mutate(free_access = ACCESS == "Free Access",
         hispanic = HISP != "Not Spanish/Hispanic/Latino",
         black = RACBLK == "Yes",
         female = SEX =="Female",
         white = hispanic == FALSE & black == FALSE,
         low_education = as.integer(SCHL) <= 14,
         high_school_GED = as.integer(SCHL) > 15 & as.integer(SCHL) <= 17,
         some_college = as.integer(SCHL) > 17 & as.integer(SCHL) <= 20,
         undergrad_or_higher = as.integer(SCHL) > 20,
         age = AGEP,
         age_square = AGEP^2
         ) 

base_model <- base_model %>% mutate(ln_earnings = log(WAGP))

model_04 <- lm(ln_earnings ~ free_access + hispanic + black + female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, data = base_model)

model_white<- lm(ln_earnings ~ free_access+female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,white == TRUE))

model_black<- lm(ln_earnings ~ free_access+female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,black == TRUE))

model_hispanic<- lm(ln_earnings ~ free_access+female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,hispanic == TRUE))
```

```{r, results='asis', echo=FALSE, error=FALSE, warning=FALSE}
stargazer(model_04, title="Results", 
          covariate.labels = c("Free Access", "Hispanic", "Black",
                               "Female", "Age (Years)", "Age Squared", "High School Diploma or GED",
                               "Some College", "Undergraduate Degree or Higher"),
          header = FALSE)

```

```{r, results='asis', echo=FALSE, error=FALSE, warning=FALSE}
stargazer(model_white, model_black, model_hispanic, 
          title="Results", 
          covariate.labels = c("Free Access", "Female", "Age (Years)", "Age Squared", 
                               "High School Diploma or GED", "Some College", 
                               "Undergraduate Degree or Higher"),
          column.labels = c("White", "Black", "Hispanic"),
          header = FALSE)
```






