---
title: "The Impact of Technology Access on Earnigs"
author: "Group 06"
date: "January 28, 2019"
output:
  html_document: 
    toc: true
    toc_depth: 2
    toc_float: true
  pdf_document: default
  word_document: default
---
# Step_0: Research Question and Variable Selection

__Research Questions:__ \n

With a past of racial inequality in the southern states it would be interesting to examine how race/ethnicity is related to access to the internet and other technologies and how not having or having access could effect a persons earnings. This can be explored via the following research questions:\n

  1. In the state of Alabama, how to earnings vary by access to the internet/technology?
  2. Does the internet/technology premium vary by race and ethnicity?

__Variable Selection__:\n

To answer the research questions, this study looks at the variables which will reveal the demographic characteristics - such as "age" or "race"; socioeconomic status - such as "employment status" or "wage"; and the access to techonlogy - such as "access to internet."\n

- The following variables will be included in this study. (Some other varialbes will be used as connector when joining tables but will not be included in the final dataframe.) \n

  1. AGEP # Age: 0#99
  2. SEMP # Self-employment income past 12 months (signed)
  3. WAGP # Wages or salary income past 12 months
  4. HISP # Recoded detailed Hispanic origin
  5. NATIVITY # Nativity
  6. PERNP # Total person's earnings
  7. RAC1P # Recoded detailed race code
  8. RACAIAN # American Indian and Alaska Native recode
  9. RACASN # Asian recode
  10. RACBLK # Black or African American recode
  11. RACNH # Native Hawaiian recode
  12. RACWHT # White recode
  13. WAOB # World area of birth
  14. WKW #Weeks worked during past 12 months
  15. ACCESS # Access to the Internet


# Step_1: System Setup
Before starting the program we need to clean the environment and import the library.\n

The loaded librariesin the code are:\n

  1. tidyverse
  2. here 
  3. DataExplorer
  
```{r setup, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Clear environment
rm(list = ls(all = TRUE))

# check if package is installed, if not installed then load
if(!require(here)){
  install.packages("here")
}
library(here)

if(!require(tidyverse)){
  install.packages("tidyverse")
}
library(tidyverse)

if(!require(DataExplorer)){
  install.packages("DataExplorer")
}
library(DataExplorer)

source(here("code","function_convert_CSV_to_vector.R"))
```

# Step_2: Data Cleaning

## Step_2.1: Import the raw data

Here we will import the raw data of personal income (ss16pal) into the dataframe as "base."
```{r import_data, message=FALSE}
#load the persional data to variable Base
base <- read_csv(here('raw_data','ss16pal.csv'))%>% 
  select(SERIALNO, #Serial Number to link House Hold and Personal data
         AGEP,  # Age: 0#99
         SEMP,  # Self-employment income past 12 months (signed)
         WAGP,  # Wages or salary income past 12 months
         HISP,  # Recoded detailed Hispanic origin
         NATIVITY,  # Nativity
         PERNP, # Total person's earnings
         RAC1P, # Recoded detailed race code
         RACAIAN, # American Indian and Alaska Native recode
         RACASN,  # Asian recode
         RACBLK,  # Black or African American recode
         RACNH, # Native Hawaiian recode
         RACWHT,  # White recode
         WAOB,  # World area of birth
         WKW)   #Weeks worked during past 12 months
```

We need to import the household level information (ss16hal) about internet access and merge the data into "base."

```{r household_info, message=FALSE}
#Link the house hould data
househould <- read_csv(here('raw_data','ss16hal.csv'))%>% 
  select(SERIALNO, #Serial Number to link House Hold and Personal data
        ACCESS # Access to the Internet
  )
base <- base %>% left_join(househould) %>% select(-SERIALNO)
rm(househould)
```

## Step_2.2: Structure of raw data
```{r first_look}
str(base)
```
Accoring to the first observation, some ordinary data will be converted into numerical type, and some categorical data will be converted into factor type. 


## Step_2.3: Convert the variable types

Convert the categorical data into factor type and numetical data into numeric type. The factorial data in this dataframe will keep its representative code instead of the category name to save the memory space. 
```{r converting_data, message=FALSE}
#conver the catagorical variable type
base <- base %>% mutate(HISP= as.factor(HISP),
                        RAC1P= as.factor(RAC1P),
                        RACAIAN = as.factor(RACAIAN),
                        RACASN = as.factor(RACASN),
                        RACBLK = as.factor(RACBLK),
                        RACNH = as.factor(RACNH),
                        RACWHT = as.factor(RACWHT),
                        WAOB = as.factor(WAOB),
                        WKW = as.factor(WKW),
                        ACCESS = as.factor(ACCESS),
                        NATIVITY = as.factor(NATIVITY)
                        )

#conver the numeric variable type
base <- base %>% mutate(AGEP = as.numeric(AGEP),
                        SEMP= as.numeric(SEMP),
                        WAGP= as.numeric(WAGP),
                        PERNP= as.numeric(PERNP)
                        )
```

Replace the factor level to the choices in the survery question for easy understanding.
```{r, message=FALSE}
levels(base$ACCESS) <- convertCSV2Factor("ACCESS")
levels(base$HISP) <- convertCSV2Factor("HISP")
levels(base$RAC1P) <- convertCSV2Factor("RAC1P")
levels(base$RACAIAN) <- convertCSV2Factor("RACAIAN")
levels(base$RACASN) <- convertCSV2Factor("RACASN")
levels(base$RACBLK) <- convertCSV2Factor("RACBLK")
levels(base$RACNH) <- convertCSV2Factor("RACNH")
levels(base$RACWHT) <- convertCSV2Factor("RACWHT")
levels(base$WKW) <- convertCSV2Factor("WKW")
levels(base$WAOB) <- convertCSV2Factor("WAOB")
levels(base$NATIVITY) <- convertCSV2Factor("NATIVITY")
```
==================

# Step_3: EDA

## Step_3.1 Brief of data structure
```{r echo=FALSE}
str(base)
```

## Step_3.2 Brief of data summary
```{r echo=FALSE}
summary(base)
```

## Step_3.3 Missing data
```{r echo=FALSE}
plot_missing(base)
```
The variable of "WKW", Weeks worked during past 12 months, has a great amount of missing data.

## Step_3.4 Brief of "Age Dispersion"

```{r eda_age}
base %>% select(AGEP) %>%  summary()
base %>% ggplot(aes(AGEP)) + geom_histogram(binwidth = 3, fill = "white", colour = "black") + xlab("Age") + ylab("Counts") + ggtitle("Age Dispersion") + theme(plot.title = element_text(hjust = 0.5))
```

## Step_3.5 Brief of "Wage"

- Summary of wages 
```{r}
summary(base$WAGP)
```

- Box plot of wages
```{r warning=FALSE}
ggplot(data = base, aes(y = WAGP)) + geom_boxplot()  + coord_flip() + ylab("Wage") + ggtitle("Wage - Box Plot") + theme(plot.title = element_text(hjust = 0.5))
```
- Density of wages
```{r warning=FALSE}
base  %>% ggplot(aes(x = WAGP)) + geom_density() + xlab("Wage") +ylab("Density") + ggtitle("Wage - Density") + theme(plot.title = element_text(hjust = 0.5))
```

- Box plot of wages by races
```{r warning=FALSE}
ggplot(data = base, aes(x = RAC1P, y = WAGP)) + geom_boxplot() + coord_flip() + xlab("Race") + ylab("Wage") + ggtitle("Wage by Races - Box Plot") + theme(plot.title = element_text(hjust = 0.5))
```

- Density of wages by races
```{r warning=FALSE}
ggplot(data = base, aes(x = WAGP, fill = RAC1P)) + geom_density(alpha = .2) + xlab("Wage") + ylab("Density") + ggtitle("Wage by Races - Density") + theme(plot.title = element_text(hjust = 0.5))
```

__Note:__ \n

category in the graph above | corresponding title in the survey
----------------------------|----------------------------------
Two more | Two or More Races
Others | Some Other Race alone
Hawaiian | Native Hawaiian and Other Pacific Islander alone
Asian | Asian alone
Tribes or Others | American Indian and Alaska Native tribes specified; or American Indian or Alaska Native, not specified and no other races
American Indian | American Indian alone
African American | Black or African American alone
White | White alone

## Step_3.6 Brief of "Access to Internet"

- Summary of access to internet
```{r}
summary(base$ACCESS)

```
```{r}
ggplot(data = base, aes(x = ACCESS)) + geom_bar() + xlab("") + ylab("Sum of Access to Internet") + coord_flip() + ggtitle("Summary of Access to Internet") + theme(plot.title = element_text(hjust = 0.5))
```

- Summary of access to internet by races
```{r}
base2 <- base %>% 
        select(c("RAC1P","ACCESS")) %>% 
        group_by(RAC1P, ACCESS) %>% 
        summarise(counts = n())
```

```{r}
ggplot(data = base2, aes(x = ACCESS, y = counts, fill = RAC1P)) + geom_col(position = "dodge") + coord_flip() + xlab("") + ylab("Sum of Access to Internet")  + ggtitle("Summary of Access to Internet") + theme(plot.title = element_text(hjust = 0.5)) + labs(fill = "Race")
rm("base2")
```

# Step_4: Criteria for filtering the data
__1. Tolerance for missing data: __

The tolerance for missing date varies by research topic. The tolerance for missing data is arbitory in this study is arbitrarily set at 50% - __the missing records in a variable should not be more than half of the total observations (Data Filtering Criteria 1)__.   

Because the variable of __"WKW"__, weeks worked during past 12 months, has about 53.34% of missing data over all observation, this variable will be __REMOVED__. 

```{r}
base$WKW <- NULL
```
__2. Age range: __

The ages of cases covered in the survey are ranging from __0 year old to 93 years old__. There is __no missing value__ under the age category.

According to the regulation from US Department of Labor, the legal working age is 14 (for most non-agricultural work), and the early retirement age is 62. This study will looks at the __age group of 14~62 years old (Data Filtering Criteria 2)__.  

```{r}
base <- base %>% 
        mutate( age_group = case_when(
                            AGEP < 14 ~ "Other ages (exclude)",
                            AGEP >= 14 & AGEP <= 62 ~ "Age 14-62 (include)",
                            AGEP > 62 ~ "Other ages (exclude)"
        ))
base$age_group <- as.factor(base$age_group)
```

```{r}
base %>% ggplot(aes(AGEP)) + geom_histogram(binwidth = 3, fill = "white", colour = "black") + xlab("Age") + ylab("Counts") + ggtitle("Age Dispersion\n This study focuses on the group in the age between 14 to 62") + theme(plot.title = element_text(hjust = 0.5)) + facet_grid(age_group ~ .)

```

Remove the records which are out of the age group of 14 to 62.  

```{r eda_age_filter}
base <- base %>%  filter(AGEP>=14 & AGEP <= 62)
base$age_group <- NULL
```

__3. Wage: __

This study excludes the out-liers of super rich and upper level, which do not contribute to the discussion of premium from access to internet. Considering the hourly wage in US was $7.25 back to 2014 and this study considering people working more than 40 weeks as the full time employees, this minimum income for full-time emloyees would be 7.25 USD/Hour * 40 Weeks * 40 Hours/Week = 11600 USD. __Yearly wage from 11,600 up to the third quartile(54,000) will be included in this study (Data Filtering Criteria 2)__. The miss data and rests will be removed.   

```{r}
base <- base %>%  filter(WAGP<54000 & WAGP > 11600)
```

# Step_5: Data summary - After the restriction
```{r }
summary(base)
```


==================

# Step_6: Examine access to the internet and its relation to earnings.

```{r}
base %>% select(ACCESS) %>% summary()

base %>%  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>% 
  ggplot(aes(WAGP,color=ACCESS)) + geom_density() + 
  labs(title="Earnings x Access to the internet",
       x ="Earnings", y = "Density") +
  scale_color_discrete(name="Internet Access",
                      labels=c("Free Access", "No Access"))

base %>%  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>%
  ggplot() + geom_histogram(aes(x=WAGP,fill=ACCESS), binwidth = 10000) 

base %>%  
  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>% 
  group_by(ACCESS, RAC1P) %>% 
  summarise (mean(WAGP),sd(WAGP),min(WAGP),max(WAGP), median(WAGP))
```

## Step_7 Race and access to the internet

```{r}
base %>% select(RAC1P) %>% summary()
base %>% filter(ACCESS=="No Access") %>% select(RAC1P) %>% summary() #person doesn't have ACCESS to Internet

base %>%  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>%
  ggplot() + geom_bar(aes(x = RAC1P, fill=ACCESS)) + 
  coord_flip() + 
  scale_x_discrete(labels = c('White', 'Black', 'Native American', "Multiple tribes", "Asian", 'Pacific Islander', 'Other', 'Two or more Races')) +
  theme(legend.position= c(0.62,0.85))

```

__Building the model__
```{r}
base_model <- base %>%  
  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>%
  mutate(free_access = ACCESS == "Free Access",
         hispanic = HISP != "Not Spanish/Hispanic/Latino",
         black = RACBLK == "Yes"
         ) 
  
model_01 <- lm(WAGP ~ free_access, data = base_model)

model_01 %>% summary()

model_02 <- lm(WAGP ~ free_access + hispanic + black, data = base_model)
model_02 %>% summary()
```





