---
title: "The Impact of Technology Access on Earnigs"
author: "Group 06"
date: "January 28, 2019"
output:
  html_document:
    df_print: paged
---

# 0. Research Question and Variable Selection

__Research Questions:__ \n

With a past of racial inequality in the southern states it would be interesting to examine how race/ethnicity is related to access to the internet and other technologies and how not having or having access could effect a persons earnings. This can be explored via the following research questions:\n

  1. In the state of Alabama, how to earnings vary by access to the internet/technology?
  2. Does the internet/technology premium vary by race and ethnicity?

__Variable Selection__:\n

To answer the research questions, this study looks at the variables which will reveal the demographic characteristics - such as "age" or "race"; socioeconomic status - such as "employment status" or "wage"; and the access to techonlogy - such as "access to internet."\n

- The following variables will be included in this study. (Some other varialbes will be used as connector when joining tables but will not be included in the final dataframe.) \n

  1. AGEP # Age: 0#99
  2. SEMP # Self-employment income past 12 months (signed)
  3. WAGP # Wages or salary income past 12 months
  4. HISP # Recoded detailed Hispanic origin
  5. NATIVITY # Nativity
  6. PERNP # Total person's earnings
  7. RAC1P # Recoded detailed race code
  8. RACAIAN # American Indian and Alaska Native recode
  9. RACASN # Asian recode
  10. RACBLK # Black or African American recode
  11. RACNH # Native Hawaiian recode
  12. RACWHT # White recode
  13. SCHL # Educational attainment
  13. WAOB # World area of birth
  14. WKW #Weeks worked during past 12 months
  15. ACCESS # Access to the Internet


# 1. System Setup
The following chunk of code (hidden from this html file) are applied to clean the working environment. There libraries and  one external function are loaded.\n

Three  loaded libraries are:\n

  1. tidyverse
  2. here 
  3. DataExplorer

One external function was programmed by the team. It is located at folder of "code", and the function is:\n

  1. function_convert_CSV_to_vector.R 
  
```{r setup, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
# Clear environment
rm(list = ls(all = TRUE))

# check if package is installed, if not installed then load
if(!require(here)){
  install.packages("here")
}
library(here)

if(!require(tidyverse)){
  install.packages("tidyverse")
}
library(tidyverse)

if(!require(DataExplorer)){
  install.packages("DataExplorer")
}
library(DataExplorer)

if(!require(lmtest)){
  install.packages("lmtest")
}
library(lmtest)

if(!require(jtools)){
  install.packages("jtools")
}
library(jtools)

source(here("code","function_convert_CSV_to_vector.R"))
```

# 2. Data Cleaning

## 2.1 Import the raw data

Import the raw data of personal income (ss16pal) into the dataframe of "base."
```{r import_data, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
#load the persional data to variable Base
base <- read_csv(here('raw_data','ss16pal.csv'))%>% 
 dplyr::select(SERIALNO, #Serial Number to link House Hold and Personal data
         AGEP,  # Age: 0#99
         SEMP,  # Self-employment income past 12 months (signed)
         WAGP,  # Wages or salary income past 12 months
         HISP,  # Recoded detailed Hispanic origin
         NATIVITY,  # Nativity
         PERNP, # Total person's earnings
         RAC1P, # Recoded detailed race code
         RACAIAN, # American Indian and Alaska Native recode
         RACASN,  # Asian recode
         RACBLK,  # Black or African American recode
         RACNH, # Native Hawaiian recode
         RACWHT,  # White recode
         SCHL,   # Educational attainment
         SEX,     # SEX
         WAOB,  # World area of birth
         WKW)   #Weeks worked during past 12 months
```

Import the household level information (ss16hal) about internet access and merge the data into "base."
```{r household_info, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
#Link the house hould data
househould <- read_csv(here('raw_data','ss16hal.csv'))%>% 
  dplyr::select(SERIALNO, #Serial Number to link House Hold and Personal data
        ACCESS # Access to the Internet
  )
base <- base %>% left_join(househould) %>% dplyr::select(-SERIALNO)
rm(househould)
```

## 2.2 Structure of raw data
Based on the first observation below, some ordinary data will be converted into numerical type, and some categorical data will be converted into factor type. 
```{r first_look, echo = FALSE}
str(base)
```

## 2.3 Convert the variable types
Convert the categorical data into factor type and numetical data into numeric type. The factorial data in this dataframe will keep its representative code instead of the category name to save the memory space. 
```{r converting_data, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
#conver the catagorical variable type
base <- base %>% mutate(HISP= as.factor(HISP),
                        RAC1P= as.factor(RAC1P),
                        RACAIAN = as.factor(RACAIAN),
                        RACASN = as.factor(RACASN),
                        RACBLK = as.factor(RACBLK),
                        RACNH = as.factor(RACNH),
                        RACWHT = as.factor(RACWHT),
                        SCHL = as.factor(SCHL),
                        SEX = as.factor(SEX),
                        WAOB = as.factor(WAOB),
                        WKW = as.factor(WKW),
                        ACCESS = as.factor(ACCESS),
                        NATIVITY = as.factor(NATIVITY)
                        )

#conver the numeric variable type
base <- base %>% mutate(AGEP = as.numeric(AGEP),
                        SEMP= as.numeric(SEMP),
                        WAGP= as.numeric(WAGP),
                        PERNP= as.numeric(PERNP)
                        )
```

Replace the factor levels to the contents in the survery question for easy understanding.
```{r, echo=TRUE, include=FALSE, message=FALSE, error=FALSE, warning=FALSE}
levels(base$ACCESS) <- convertCSV2Factor("ACCESS")
levels(base$HISP) <- convertCSV2Factor("HISP")
levels(base$RAC1P) <- convertCSV2Factor("RAC1P")
levels(base$RACAIAN) <- convertCSV2Factor("RACAIAN")
levels(base$RACASN) <- convertCSV2Factor("RACASN")
levels(base$RACBLK) <- convertCSV2Factor("RACBLK")
levels(base$RACNH) <- convertCSV2Factor("RACNH")
levels(base$RACWHT) <- convertCSV2Factor("RACWHT")
levels(base$SCHL) <- convertCSV2Factor("SCHL")
levels(base$SEX) <- convertCSV2Factor("SEX")
levels(base$WKW) <- convertCSV2Factor("WKW")
levels(base$WAOB) <- convertCSV2Factor("WAOB")
levels(base$NATIVITY) <- convertCSV2Factor("NATIVITY")
```
==================


# 3. Criteria for filtering the data

__1. Age range: __

The ages of cases covered in the survey are ranging from __0 year old to 93 years old__. There is __no missing value__ under the age category.

According to the regulation from US Department of Labor, the legal working age is 14 (for most non-agricultural work), and the early retirement age is 62. This study will looks at the __age group of 14~62 years old (Data Filtering Criteria 1)__.  

```{r}
base <- base %>% 
        mutate( age_group = case_when(
                            AGEP < 14 ~ "Other ages (exclude)",
                            AGEP >= 14 & AGEP <= 62 ~ "Age 14-62 (include)",
                            AGEP > 62 ~ "Other ages (exclude)"
        ))
base$age_group <- as.factor(base$age_group)
```


Remove the records which are out of the age group of 14 to 62.  

```{r eda_age_filter}
base <- base %>%  filter(AGEP>=14 & AGEP <= 62)
base$age_group <- NULL
```

__2. Earnings: __

In this study we will consider everyone that had positive earnings in the last 12 months.   

```{r}
base <- base %>%  filter(WAGP > 0)
```

__3. Annually Working Weeks: __

The selection of full time employee is arbitrary. This study includes groups which __work more than 40 weeks a years (Data Filtering Criteria 4)__. (Please be noted: More studies shold be considered in decideing the threshold separating full-time and part-time employee. This study set the 40 week/year as the threshold for now.)

```{r}
base <- base %>% 
         filter(WKW %in% c("50 to 52 weeks worked during past 12 months", "48 to 49 weeks worked during past 12 months", "40 to 47 weeks worked during past 12 months"))
```

==================

# 4. Building Models

## 4.1 Examine access to the internet and its relation to earnings.

```{r}
base %>% dplyr::select(ACCESS) %>% summary()

base %>%  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>% 
  ggplot(aes(WAGP,color=ACCESS)) + geom_density() + 
  labs(title="Earnings x Access to the internet",
       x ="Earnings", y = "Density") +
  scale_color_discrete(name="Internet Access",
                      labels=c("Free Access", "No Access"))

base %>%  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>%
  ggplot() + geom_histogram(aes(x=WAGP,fill=ACCESS), binwidth = 10000) 

base %>%  
  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>% 
  group_by(ACCESS, RAC1P) %>% 
  summarise (mean(WAGP),sd(WAGP),min(WAGP),max(WAGP), median(WAGP))
```

__Building the model__
```{r}
base_model <- base %>%  
  filter(!is.na(ACCESS) & ACCESS != "Paid Access") %>%
  mutate(free_access = ACCESS == "Free Access",
         hispanic = HISP != "Not Spanish/Hispanic/Latino",
         black = RACBLK == "Yes",
         female = SEX =="Female",
         white = hispanic == FALSE & black == FALSE,
         low_education = as.integer(SCHL) <= 14,
         high_school_GED = as.integer(SCHL) > 15 & as.integer(SCHL) <= 17,
         some_college = as.integer(SCHL) > 17 & as.integer(SCHL) <= 20,
         undergrad_or_higher = as.integer(SCHL) > 20,
         age = AGEP,
         age_square = AGEP^2
         ) 
model_01 <- lm(WAGP ~ free_access, data = base_model)

model_01 %>% summary()

model_02 <- lm(WAGP ~ free_access + hispanic + black + female + 
                 high_school_GED + some_college + undergrad_or_higher, data = base_model)
model_02 %>% summary()
```
```{r}
#Conduct BPG Test & Plot Model_04 for Diagnostics
bptest(model_01)
bptest(model_02)
```

__Building the model with a log function__
```{r}
base_model <- base_model %>% mutate(ln_earnings = log(WAGP))

model_03 <- lm(ln_earnings ~ free_access, data = base_model)

model_03 %>% summary()

bptest(model_03)

model_04 <- lm(ln_earnings ~ free_access + hispanic + black + female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, data = base_model)

model_04 %>% summary()

model_05 <- lm(ln_earnings ~ hispanic + black + female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,free_access == TRUE))

model_05 %>% summary()

model_06 <- lm(ln_earnings ~ hispanic + black + female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,free_access == FALSE))

model_06 %>% summary()


model_white<- lm(ln_earnings ~ free_access+female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,white == TRUE))

model_black<- lm(ln_earnings ~ free_access+female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,black == TRUE))

model_hispanic<- lm(ln_earnings ~ free_access+female + AGEP + I(AGEP^2) +
                 high_school_GED + some_college + undergrad_or_higher, 
               data = filter(base_model,hispanic == TRUE))

#Conduct BPG Test & Plot Model_04 for Diagnostics
bptest(model_04)
par(mfrow = c(2, 2))
plot(model_04)
```

__Diagnostics of Model_04 Log Earnings__  

**Breusch_Pagan Test**  

* Since the p-value is > 0.05 we fail to reject the null hypothesis of homoskedasticity.  

**Residuals vs Fitted Plot.**  

* The line is horizonal and there is no distinct patterns, indicating a linear relationship.

**Normal Q-Q**  

* The residuals seem to somewhat follow the dashed line, therefore it appears that the residuals are normally distributed.  

**Scale-Location**  

* Since the line is close to horizontal and based on our BP test we can see that we do not have a heteroscedasticity problem with this model.

__Trying new package__ 

```{r}
export_summs(model_04,
             model.names = c("Access and Earnings"),
             coefs = c("Have free access" = "free_accessTRUE",
                      "Hispanic" = "hispanicTRUE",
                      "Black" = "blackTRUE",
                      "Woman" = "femaleTRUE",
                      "Age" = "AGEP",
                      "Age Square" = "I(AGEP^2)",
                      "High School Diploma or GED" = "high_school_GEDTRUE",
                      "Some College" = "some_collegeTRUE",
                      "Undergrad or higher" = "undergrad_or_higherTRUE")
            )
```


```{r}
export_summs(model_05,
             model_06, 
             model.names = c("No Free Access","Free Access"),
             coefs =c("Hispanic" = "hispanicTRUE",
                      "Black" = "blackTRUE",
                      "Woman" = "femaleTRUE",
                      "Age" = "AGEP",
                      "Age Square" = "I(AGEP^2)",
                      "High School Diploma or GED" = "high_school_GEDTRUE",
                      "Some College" = "some_collegeTRUE",
                      "Undergrad or higher" = "undergrad_or_higherTRUE")
             )
```

```{r}
export_summs(model_white,
             model_black, 
             model_hispanic, 
             model.names = c("White","Black","Hispanic"),
             coefs =c("Have free access" = "free_accessTRUE", 
                      "Woman" = "femaleTRUE",
                      "Age" = "AGEP",
                      "Age Square" = "I(AGEP^2)",
                      "High School Diploma or GED" = "high_school_GEDTRUE",
                      "Some College" = "some_collegeTRUE",
                      "Undergrad or higher" = "undergrad_or_higherTRUE")
             )
```


# 5. Updates for the week5

__The updates in the week4's report are: __
  
  1. 
  